\chapter*{Acknowledgements}
\thispagestyle{empty}

First, I would like to express my deepest gratitude to Marc Bui, who
accepted to supervise this thesis, for very interesting discussions, for his
valuable experience and guidance in matters both scientific and administrative.

I am so grateful to the eScripta team at EPHE and everyone at the Chair for
Digital Humanities at the University of Leipzig. Both provided an environment
were I was free to explore research ideas and their generous financial support
made my existence as a doctoral student a comfortable one. It was Maxim Romanov
through whom I had the first contact with Arabic text in Leipzig. I am also
eternally indebted to Daniel Stökl Ben Ezra and Peter Stokes for fending off
every imposition on my time during the last weeks of writing. 

I thank the OpenITI team for their incessant lobbying to get institutional
support for research into Arabic OCR. Without them, none of this would have
been possible. A special thanks goes to Matthew Thomas Miller and Jonathan
Allen who responded to my endless question on details of the Arabic script with
a celerity unheard of in academia.

The people of ALMAnaCH at INRIA should not be forgotten. They welcomed a
clueless foreigner on his arrival in France and integrated me in the lab.

I have wonderful friends who were there even at times when all they heard from
me was about this thesis. A huge thanks goes to Louise for helping improve the
French summary. Above all I thank Monika for reminding me in the darkest
moments that instant ramen is not the foundation of a balanced diet.
